---
title: "Survival Analysis in R: A Worked Beginner Tutorial"
subtitle: "Heart Failure Clinical Records"
author: "cjgunase"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  github_document:
    toc: true
    html_preview: false
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

library(dplyr)
library(ggplot2)
library(readr)
library(splines)
library(survival)

heart <- read_csv(
  "../data/processed/heart_failure_clean.csv",
  show_col_types = FALSE
) |>
  mutate(
    age_per_10_years = age / 10,
    ef_per_5_points = ejection_fraction / 5,
    creatinine_per_doubling = log2(serum_creatinine),
    sodium_per_5_units = serum_sodium / 5,
    ef_group = factor(
      if_else(ejection_fraction < 40, "EF < 40%", "EF ≥ 40%"),
      levels = c("EF < 40%", "EF ≥ 40%")
    ),
    sex_group = factor(
      sex,
      levels = c(0, 1),
      labels = c("Female", "Male")
    )
  )

data_quality <- read_csv(
  "../outputs/tables/data-quality-summary.csv",
  show_col_types = FALSE
)

km_times <- read_csv(
  "../outputs/tables/kaplan-meier-selected-times.csv",
  show_col_types = FALSE
)

grouped_estimates <- read_csv(
  "../outputs/tables/grouped-survival-selected-times.csv",
  show_col_types = FALSE
)

log_rank_results <- read_csv(
  "../outputs/tables/log-rank-tests.csv",
  show_col_types = FALSE
)

cox_results <- read_csv(
  "../outputs/tables/cox-adjusted-results.csv",
  show_col_types = FALSE
)

nonlinearity_results <- read_csv(
  "../outputs/tables/diagnostic-nonlinearity-tests.csv",
  show_col_types = FALSE
)

bootstrap_concordance <- read_csv(
  "../outputs/tables/bootstrap-concordance-summary.csv",
  show_col_types = FALSE
)

bootstrap_calibration <- read_csv(
  "../outputs/tables/bootstrap-calibration-90-days.csv",
  show_col_types = FALSE
)
```

# Purpose of this tutorial

Survival analysis studies **whether an event occurs and how long it takes**. It
is especially useful when participants have different follow-up times or when
their eventual event times are unknown.

This tutorial answers:

> Which baseline patient characteristics are associated with time to death
> among patients with heart failure?

The analysis is educational and prognostic. It does not estimate causal effects
and must not be used for clinical decisions.

## Learning goals

By the end, a reader should be able to:

1. distinguish an observed event from censoring;
2. explain a risk set and survival function;
3. calculate and interpret Kaplan--Meier survival;
4. compare curves with a log-rank test;
5. interpret adjusted Cox hazard ratios;
6. recognize proportional-hazards and functional-form assumptions;
7. distinguish apparent from internally validated performance.

# 1. The data

The [UCI Heart Failure Clinical Records
dataset](https://archive.ics.uci.edu/dataset/519/heart) contains records for 299
patients. Its license is CC BY 4.0.

```{r first-rows}
heart |>
  select(
    followup_days,
    death,
    age,
    ejection_fraction,
    serum_creatinine,
    serum_sodium,
    sex_group
  ) |>
  head() |>
  knitr::kable(
    caption = "Example analysis records"
  )
```

The survival outcome has two columns:

- `followup_days`: observed time under follow-up;
- `death`: 1 if death was observed and 0 if the patient was censored.

```{r data-quality, echo=FALSE}
data_quality |>
  mutate(
    value = if_else(
      metric == "Death proportion",
      sprintf("%.1f%%", 100 * value),
      format(round(value, 1), trim = TRUE)
    )
  ) |>
  knitr::kable(
    caption = "Data-quality and outcome summary"
  )
```

```{r marker-distributions, echo=FALSE, out.width="95%", fig.cap="Distributions of prespecified clinical predictors."}
knitr::include_graphics(
  "../outputs/figures/clinical-marker-distributions.png"
)
```

Serum creatinine and creatinine phosphokinase are strongly right-skewed.
Extreme values should be checked for plausibility, not automatically deleted.

# 2. Censoring

A patient censored after 100 days tells us:

\[
T>100
\]

The patient was known to remain event-free for at least 100 days, but their
eventual death time is unknown.

Censored patients:

- contribute to risk sets while observed;
- leave the risk set at censoring;
- are not counted as deaths;
- are not assumed to survive forever.

```{r survival-object}
survival_outcome <- with(
  heart,
  Surv(followup_days, death)
)

head(survival_outcome)
```

`Surv()` represents time and event status together. A printed `+` identifies a
censored observation.

> The central assumption is independent censoring: after accounting for
> relevant information, censoring should not be systematically related to
> unobserved future event risk.

# 3. Kaplan--Meier survival

The survival function is:

\[
S(t)=P(T>t)
\]

At event time \(t_j\):

- \(n_j\) is the number at risk immediately before the time;
- \(d_j\) is the number of deaths at the time.

The conditional probability of surviving that event time is:

\[
p_j=\frac{n_j-d_j}{n_j}
\]

Kaplan--Meier multiplies these conditional probabilities:

\[
\hat S(t)=
\prod_{t_j\leq t}
\left(1-\frac{d_j}{n_j}\right)
\]

## Calculation with the first events

At day 4, one of 299 patients died:

\[
\hat S(4)=\frac{298}{299}=0.9967
\]

At day 6, one of the remaining 298 died:

\[
\hat S(6)=
\frac{298}{299}\times\frac{297}{298}
=0.9933
\]

At day 7, two of 297 at-risk patients died:

\[
\hat S(7)=
0.9933\times\frac{295}{297}
=0.9866
\]

```{r fit-km}
overall_km <- survfit(
  Surv(followup_days, death) ~ 1,
  data = heart,
  conf.type = "log"
)
```

```{r overall-km, echo=FALSE, out.width="95%", fig.cap="Overall Kaplan--Meier estimate with confidence interval, censor marks, and numbers at risk."}
knitr::include_graphics(
  "../outputs/figures/overall-kaplan-meier.png"
)
```

```{r km-table, echo=FALSE}
km_times |>
  transmute(
    `Time, days` = time_days,
    `Number at risk` = n_risk,
    `Survival` = sprintf("%.1f%%", 100 * survival),
    `95% CI` = sprintf(
      "%.1f%%–%.1f%%",
      100 * lower_95,
      100 * upper_95
    )
  ) |>
  knitr::kable(
    caption = "Overall survival estimates"
  )
```

At 90 days, estimated survival was 76.3% (95% CI: 71.5%--81.3%).

The curve never reached 50%, so median survival was **not reached**. This does
not mean median survival is infinite; it means the available follow-up cannot
estimate it.

# 4. Comparing groups

Grouped Kaplan--Meier curves describe unadjusted survival patterns.

```{r grouped-km-fit}
ef_km <- survfit(
  Surv(followup_days, death) ~ ef_group,
  data = heart
)

ef_log_rank <- survdiff(
  Surv(followup_days, death) ~ ef_group,
  data = heart
)
```

```{r ef-km, echo=FALSE, out.width="95%", fig.cap="Kaplan--Meier estimates by educational ejection-fraction group."}
knitr::include_graphics(
  "../outputs/figures/kaplan-meier-ejection-fraction.png"
)
```

```{r grouped-table, echo=FALSE}
grouped_estimates |>
  filter(comparison == "Ejection fraction group") |>
  transmute(
    Group = group,
    `Time, days` = time_days,
    `Number at risk` = n_risk,
    `Survival` = sprintf("%.1f%%", 100 * survival),
    `95% CI` = sprintf(
      "%.1f%%–%.1f%%",
      100 * lower_95,
      100 * upper_95
    )
  ) |>
  knitr::kable(
    caption = "Survival estimates by ejection-fraction group"
  )
```

The log-rank test compares observed with expected deaths over follow-up:

```{r logrank-table, echo=FALSE}
log_rank_results |>
  transmute(
    Comparison = comparison,
    `Chi-square` = round(chi_square, 3),
    df = degrees_freedom,
    `p-value` = format.pval(p_value, digits = 3, eps = 0.001)
  ) |>
  knitr::kable(
    caption = "Log-rank tests"
  )
```

There was evidence of different survival curves between ejection-fraction
groups (\(\chi^2=9.47\), \(p=0.0021\)). There was no evidence of a sex difference
(\(p=0.950\)).

These are unadjusted associations. A p-value does not measure effect size and
does not establish causation.

# 5. Cox proportional-hazards regression

The Cox model is:

\[
h(t\mid X)
=
h_0(t)\exp(\beta_1X_1+\cdots+\beta_pX_p)
\]

At every death time, the partial likelihood compares the patient who died with
patients still in the risk set. The model estimates relative hazard without
requiring a parametric shape for \(h_0(t)\).

```{r cox-fit}
adjusted_model <- coxph(
  Surv(followup_days, death) ~
    age_per_10_years +
    ef_per_5_points +
    creatinine_per_doubling +
    sodium_per_5_units +
    sex_group,
  data = heart,
  ties = "efron"
)
```

```{r cox-table, echo=FALSE}
cox_results |>
  transmute(
    Predictor = predictor,
    HR = sprintf("%.2f", hazard_ratio),
    `95% CI` = sprintf("%.2f–%.2f", lower_95, upper_95),
    `p-value` = format.pval(p_value, digits = 3, eps = 0.001)
  ) |>
  knitr::kable(
    caption = "Adjusted Cox proportional-hazards results"
  )
```

```{r forest-plot, echo=FALSE, out.width="95%", fig.cap="Adjusted hazard ratios from the initial linear Cox model."}
knitr::include_graphics(
  "../outputs/figures/cox-adjusted-forest-plot.png"
)
```

Interpretation examples:

- A 10-year age increase was associated with 52% higher mortality hazard.
- A 5-point ejection-fraction increase was associated with 19% lower hazard.
- A creatinine doubling was associated with 85% higher hazard.

Each interpretation is conditional on the other included predictors. Hazard
ratios are relative rates, not direct differences in survival probability.

# 6. Model diagnostics

## Proportional hazards

The Cox model assumes each hazard ratio is approximately constant over
follow-up.

```{r ph-test}
ph_assessment <- cox.zph(adjusted_model)
ph_assessment
```

The global test gave \(p=0.233\), providing no clear evidence of a violation.
Ejection fraction had the strongest possible trend (\(p=0.060\)), so graphical
assessment remained important.

```{r schoenfeld, echo=FALSE, out.width="95%", fig.cap="Centered scaled Schoenfeld diagnostics."}
knitr::include_graphics(
  "../outputs/figures/cox-schoenfeld-residuals.png"
)
```

## Nonlinear relationships

The initial Cox model assumes a straight-line relationship between each
continuous predictor and log hazard. Natural splines test whether that shape is
too restrictive.

```{r nonlinearity-table, echo=FALSE}
nonlinearity_results |>
  transmute(
    Predictor = predictor,
    `Added df` = added_degrees_freedom,
    `LR chi-square` = round(likelihood_ratio_chi_square, 2),
    `p-value` = format.pval(p_value, digits = 3, eps = 0.001),
    `Linear AIC` = round(linear_model_aic, 1),
    `Spline AIC` = round(predictor_spline_model_aic, 1)
  ) |>
  knitr::kable(
    caption = "Predictor-specific functional-form tests"
  )
```

Ejection fraction showed strong evidence of nonlinearity (\(p=0.0013\)). The
estimated hazard increased sharply at very low values and was comparatively
flat around and above 40%.

```{r ef-spline, echo=FALSE, out.width="95%", fig.cap="Adjusted nonlinear ejection-fraction association, referenced to EF 40%."}
knitr::include_graphics(
  "../outputs/figures/cox-ejection-fraction-spline.png"
)
```

Confidence intervals widen at sparse extremes. The rug marks show where
patients were observed.

## Unusual and influential observations

One patient had an absolute deviance residual greater than 3. DFBETAS also
identified observations that affected individual coefficients.

```{r influence-plots, echo=FALSE, out.width="95%", fig.cap="Deviance-residual and DFBETAS screening."}
knitr::include_graphics(
  c(
    "../outputs/figures/cox-deviance-residuals.png",
    "../outputs/figures/cox-dfbetas.png"
  )
)
```

These observations are not automatically errors. They may represent
scientifically informative combinations of predictors and outcomes.

# 7. Bootstrap internal validation

The refined model used a 3-df ejection-fraction spline. It was fixed before
validation.

For each of 500 bootstrap resamples:

1. sample patients with replacement;
2. refit the complete model;
3. calculate concordance in the bootstrap sample;
4. apply the model to the original sample;
5. estimate optimism as training minus test performance;
6. generate predictions for out-of-bag patients.

```{r bootstrap-summary, echo=FALSE}
bootstrap_concordance |>
  filter(
    statistic %in% c(
      "Requested bootstrap resamples",
      "Successful bootstrap fits",
      "Apparent concordance",
      "Mean estimated optimism",
      "Optimism-corrected concordance"
    )
  ) |>
  mutate(
    value = if_else(
      grepl("resamples|fits", statistic),
      format(round(value), trim = TRUE),
      sprintf("%.3f", value)
    )
  ) |>
  knitr::kable(
    caption = "Bootstrap internal-validation summary"
  )
```

```{r bootstrap-c, echo=FALSE, out.width="95%", fig.cap="Bootstrap validation of concordance."}
knitr::include_graphics(
  "../outputs/figures/bootstrap-concordance.png"
)
```

Apparent concordance was 0.750. After subtracting estimated optimism of 0.016,
corrected concordance was 0.735.

## Calibration

Discrimination asks whether the model ranks patients correctly. Calibration
asks whether predicted probabilities agree with observed survival.

```{r calibration, echo=FALSE, out.width="95%", fig.cap="Apparent and out-of-bag calibration of 90-day survival."}
knitr::include_graphics(
  "../outputs/figures/bootstrap-calibration-90-days.png"
)
```

Out-of-bag calibration was broadly reasonable, although the second prediction
quintile overpredicted survival by approximately eight percentage points.
Groups contained only about 60 patients, producing wide confidence intervals.

# 8. Conclusions

Within this dataset:

1. survival declined progressively over observed follow-up;
2. very low ejection fraction was associated with poorer survival;
3. older age and higher serum creatinine were associated with higher adjusted
   mortality hazard;
4. the ejection-fraction association was nonlinear;
5. the refined model had optimism-corrected concordance around 0.735;
6. calibration was imperfect in part of the risk range.

These findings describe associations in a small observational dataset. They do
not show causal effects or establish performance in other populations.

# 9. Reproduce the project

Run the scripts from the repository root:

```{bash, eval=FALSE}
Rscript R/01-download-data.R
Rscript R/02-exploratory-analysis.R
Rscript R/03-kaplan-meier.R
Rscript R/04-group-comparisons.R
Rscript R/05-cox-regression.R
Rscript R/06-model-diagnostics.R
Rscript R/07-bootstrap-validation.R
```

Then render this report:

```{r render-report, eval=FALSE}
rmarkdown::render(
  "report/survival-analysis-tutorial.Rmd",
  output_format = "all"
)
```

The detailed living study guide is available in `notes.md`.

# 10. Limitations

- The sample contains only 299 patients and 96 observed deaths.
- Predictor selection and transformations depend partly on this dataset.
- Group cutpoints lose continuous information.
- Internal bootstrap validation does not replace external validation.
- Measurement error and unmeasured predictors may affect results.
- Independent censoring cannot be fully verified from the observed data.
- The analysis cannot support causal or clinical treatment claims.
